---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 200,
  dev = "ragg_png"
)
```

# {msm.stacked}

<!-- badges: start -->
<!-- badges: end -->

The {msm.stacked} package can be used to simplify the calculation of state transition probabilities over time and the creation of stacked probabilities plot from multi-state model fits from the [{msm}](https://CRAN.R-project.org/package=msm) package.

# Installation

You can install the development version of {msm.stacked} from [GitHub](https://github.com) with:

```r
# install.packages("devtools")
devtools::install_github("RedDoorAnalytics/msm.stacked")
```

# Example

To illustrate the functionality of {msm.stacked}, we build upon the documentation for the `msm::msm()` function.
We will be using the heart transplant data:

```{r data}
library(msm)
head(cav)
```

Further details on this example dataset are included in the [vignette of the {msm} package](https://CRAN.R-project.org/package=msm/vignettes/msm-manual.pdf).

We start with a matrix of possible transitions:

```{r twoway}
twoway4.q <- rbind(
  c(-0.5, 0.25, 0, 0.25),
  c(0.166, -0.498, 0.166, 0.166),
  c(0, 0.25, -0.5, 0.25),
  c(0, 0, 0, 0)
)
twoway4.q
```

This is then used to provide starting values for the model without any additional covariate:

```{r fit}
cav.msm <- msm(
  formula = state ~ years,
  subject = PTNUM,
  data = cav,
  qmatrix = twoway4.q,
  deathexact = 4
)
cav.msm
```

We can use the `plot.msm()` function to plot survival curves from every transient state to the final, absorbing state (e.g., a state denoting death).
This is denoted in the `cav` dataset by _state 4_:

```{r plot.msm}
plot(cav.msm, from = 1:3, to = 4)
```

The {msm} package also provides functionality to calculate state transition probabilities at a given point in time.
Say we are interested in estimating the probability of being in a given state, from each state, five years after baseline; we can use the `pmatrix.msm()` function to obtain just that:

```{r pmatrix.msm}
pmatrix.msm(x = cav.msm, t = 5)
```

This shows that, for instance, study participants in State 1 at time zero have (approximately) a 52% probability of still being in State 5 after years, 14% probability of being in State 2, 9% probability of being in State 3, and 25% probability of being in State 4.

We can repeatedly call the `pmatrix.msm()` function to obtain probabilities over time, but that's a bit tedious. 
This is where the {msm.stacked} package comes in handy.

Specifically, we can use the `stacked.data.msm()` function to calculate transition probabilities over time, say, at 1 to 5 years:

```{r stacked.data.all}
library(msm.stacked)
sdd <- stacked.data.msm(model = cav.msm, tstart = 0, tforward = 5, tseqn = 6)
str(sdd)
```

This returns a tidy dataset with all transition probabilities, from and to every state, over `tseqn = 6` equally-spaced time intervals between time zero and time five. 
Focussing on transitions from State 1 only:

```{r stacked.data.from1}
subset(sdd, sdd$from == "State 1")
```

Here we see, for instance, that the probability of still being in State 1, starting from State 1, is 85% after one year, 74% after two years, 65% after three years, 58% after four years, and 52% after five years:

```{r stacked.data.from1.to1}
subset(sdd, sdd$from == "State 1" & sdd$to == "State 1")
```

The package also provides functionality to automatically produce stacked probabilities plots, for transition probabilities from and to every state.
This is implemented in the `stacked.plot.msm()` function:

```{r stacked.plot}
stacked.plot.msm(model = cav.msm, tstart = 0, tforward = 5)
```

This relies on [{ggplot2}](https://ggplot2.tidyverse.org) functionality and returns a standard `ggplot` object, which can of course be further customised beyond the default settings:

```{r stacked.plot.custom}
library(ggplot2)

stacked.plot.msm(model = cav.msm, tstart = 0, tforward = 5) +
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  labs(fill = "To:")
```

# Model with Covariates

We can of course incorporate covariates in a multi-state model and obtain predictions for a specific covariates pattern; let's demonstrate this by incorporating sex in the model above.
First, we fit a second model:

```{r fit.cov}
cav.msm.cov <- msm(
  formula = state ~ years,
  subject = PTNUM,
  data = cav,
  covariates = ~sex,
  qmatrix = twoway4.q,
  deathexact = 4
)
cav.msm.cov
```

Then, we can use the same functionality as before to obtain stacked probabilities plots:

```{r stacked.plot.cov}
stacked.plot.msm(model = cav.msm.cov, tstart = 0, tforward = 5) +
  labs(title = "Predictions for average covariates")
```

By default, this will set all covariates to their average value (as in `pmatrix.msm()`); we can, however, pass specific covariates patterns that we want to predict for:

```{r stacked.plot.cov.sex0}
stacked.plot.msm(model = cav.msm.cov, tstart = 0, tforward = 5, covariates = list(sex = 0)) +
  labs(title = "Predictions for 'sex = 0'")
```

```{r stacked.plot.cov.sex1}
stacked.plot.msm(model = cav.msm.cov, tstart = 0, tforward = 5, covariates = list(sex = 1)) +
  labs(title = "Predictions for 'sex = 1'")
```

This way we can provide clinically meaningful predictions that highlight the effect of covariates of interest on state occupancy probabilities over time.

# Model with Piecewise-Constant Intensities

By default, the {msm} package assumes constant (i.e., exponential) baseline transition intensities.
This means that predictions at t years will be the same, irrespectively of when the starting point is:

```{r stacked.plot.two.times}
stacked.plot.msm(model = cav.msm, tstart = 0, tforward = 3)
stacked.plot.msm(model = cav.msm, tstart = 5, tforward = 3, start0 = FALSE)
```

We can relax this assumption by allowing piecewise-constant baseline transition rates.
This can be done by setting the `pci` argument of `msm()`:

```{r fit.pw}
cav.msm.pw <- msm(
  formula = state ~ years,
  subject = PTNUM,
  data = cav,
  qmatrix = twoway4.q,
  deathexact = 4,
  pci = quantile(x = cav$years, probs = c(0.25, 0.50, 0.75))
)
cav.msm.pw
```

Specifically, here we set cut-points at quartiles of the observed distribution of (possibly censored) transition times.
We ignore the warning about _non-convergence_ for now - in practice, we should investigate this and try a different optimiser or "consider tightening the tolerance criteria for convergence" (according to the documentation of `msm()`).

First, we can do a likelihood ratio test to check whether the model with piecewise-constant intensities fits the data better:

```{r lrt}
lrtest.msm(cav.msm, cav.msm.pw)
```

The test is statistically significant at any usual level, thus the more flexible model seems appropriate.
Predictions of transition probabilities will now depend on the starting point `tstart`, even though `tforward` is the same:

```{r stacked.plot.two.times.pw}
stacked.plot.msm(model = cav.msm.pw, tstart = 0, tforward = 3)
stacked.plot.msm(model = cav.msm.pw, tstart = 5, tforward = 3, start0 = FALSE)
```

As expected, we see that the predicted probabilities between 0 and 3 units of time are now different compared to those between 5 and 8, given the (now) non-constant baseline intensities. 

# Excluding States

The {msm.stacked} package also include functionality to calculate (and plot) transition probabilities from only certain states of interest.
For instance, the models we fit in the previous examples includes an absorbing state, _State 4_, from which there will be no transitions.
For this example, we will be using the model with constant baseline transition rates (`cav.msm`).

Let's start with a utility function to determine the names of the state of a {msm} model fit.
This is called `states.msm()`:

```{r states}
states.msm(cav.msm)
```

Now we have the correct names used by {msm} to define each state.
Let's calculate transitions probabilities from all states, but excluding the absorbing _State 4_:

```{r transitions.no.absorbing}
stacked.data.msm(model = cav.msm, tstart = 0, tforward = 1, tseqn = 3, exclude = "State 4")
```

As you can see, transitions from the state passed to `exclude` are not reported.
We can also exclude more than one state, for instance if we want to calculate only transitions from _State 1_:

```{r transitions.only.one}
stacked.data.msm(model = cav.msm, tstart = 0, tforward = 1, tseqn = 3, exclude = c("State 2", "State 3", "State 4"))
```

Of course, this functionality is also included in the plotting function:

```{r plot.no.absorbing}
stacked.plot.msm(model = cav.msm, tstart = 0, tforward = 10, exclude = "State 4") +
  theme(legend.position = "bottom")
```
