---
title: "Confidence Intervals for Transition Probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Confidence Intervals for Transition Probabilities}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  fig.width = 6,
  fig.height = 4,
  dpi = 200
)
```

As of version 0.0.2, the {msm.stacked} package can produce confidence intervals for the transition probabilities over time as computed via the `stacked.data.msm()` function.
This has been revamped and improved in version 0.0.3.

To illustrate this functionality, we replicate the example from the introductory vignette:

```{r replicate_example}
library(msm)

twoway4.q <- rbind(
  c(-0.5, 0.25, 0, 0.25),
  c(0.166, -0.498, 0.166, 0.166),
  c(0, 0.25, -0.5, 0.25),
  c(0, 0, 0, 0)
)
twoway4.q

cav.msm <- msm(
  formula = state ~ years,
  subject = PTNUM,
  data = cav,
  qmatrix = twoway4.q,
  deathexact = 4
)
cav.msm
```

If needed, check that vignette with the following R code:

```r
vignette("A-introduction", package = "msm.stacked")
```

As demonstrated there, we can use the `stacked.data.msm()` function to calculate transition probabilities over time, say, at 1 to 5 years:

```{r stacked.data}
library(msm.stacked)

sdd <- stacked.data.msm(model = cav.msm, tstart = 0, tforward = 5, tseqn = 20)
str(sdd)
```

With this data we could, for instance, plot this over time:

```{r stacked.data.plot}
library(ggplot2)

ggplot(sdd, aes(x = t, y = p)) +
  geom_line(aes(color = to)) +
  facet_wrap(~from)
```

In the plot above, we can see transition probabilities over time from and to each state.
We might however be interested in including pointwise confidence intervals, over time, to visualise uncertainty around our estimates.

This can now be easily done by setting the `conf.int` argument to `TRUE`:

```{r stacked.data.ci}
sdd.ci <- stacked.data.msm(model = cav.msm, tstart = 0, tforward = 5, tseqn = 20, conf.int = TRUE, progress = FALSE)
str(sdd.ci)
```

Note that we set the additional argument `progress = FALSE` to omit displaying a progress bar, and that now two additional columns are available, `conf.low` and `conf.high` for the lower and upper bound of pointwise confidence intervals, respectively.

The method implemented in `stacked.data.msm()` is based on that implemented in {msm}'s `pmatrix.msm()`, but ensuring that each single replicate is used across all time points that we predict for. 
It is a parametric bootstrap approach with number of repetitions defined by the `B` argument of `stacked.data.msm()` (defaulting to 1,000) and calculating confidence intervals with the percentile method.
According to the documentation of {msm}, this approach is based on "[...] simulating `B` random vectors from the asymptotic multivariate normal distribution implied by the maximum likelihood estimates (and covariance matrix) of the log transition intensities and covariate effects, then calculating the resulting transition probability matrix for each replicate.".
This is discussed in more detail in the paper by Mandel (2013, see references below).

After calculating the confidence intervals, we can include those in a plot:

```{r stacked.data.plot.ci}
ggplot(sdd.ci, aes(x = t, y = p)) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high, fill = to), alpha = 0.2) +
  geom_line(aes(color = to)) +
  facet_wrap(~from)
```

...and there you have it.

Finally, note that this functionality supports different confidence levels (via the `alpha` argument, which is passed down to `msm::pmatrix.msm()`), multi-state models with covariates, and everything else that is supported by `stacked.data.msm()` and `msm::pmatrix.msm()` and passed down via the `...` argument.

# References

* Mandel, M. (2013). "Simulation based confidence intervals for functions with complicated derivatives." The American Statistician 67(2):76-81
